**Machine learning classification using self-reference-based OTU clustering**

### 2023-04-13 submission/response_to_reviewer.Rmd

---
output: pdf_document
geometry: margin=1.0in
font-size: 11pt
header-includes:
  - \usepackage{helvet}
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage[left]{lineno}
  - \usepackage{multirow}
  - \usepackage[none]{hyphenat}
---

## Response to Reviewer Comments

Machine learning classification using self-reference-based OTU clustering

Courtney R. Armour, Kelly L. Sovacool, William L. Close, Begüm D. Topçuoğlu, Jenna Wiens, Patrick D. Schloss

**Reviewer #1:**

Armour describe using the closed-reference OTU method OptiRef to predict screen-relevant-neoplasia (SRN) from 16S sequencing data of the fecal microbiome that was originally processed into OTUs by the de novo method OptiClust.

Is this a case study, or something more general? The results here are based on a single dataset, a single set of methods, and a single set of evaluation metrics. Thus, the generalizability of these results is unclear. This concern is exacerbated by the various places in the manuscript that infer generalizable conclusions.

\textcolor{blue}{The purpose of this paper is to provide an example of a scenario when the previously publish OptiFit algorithm could be useful. We are demonstrating here that it is possible to use your own data as a reference for consistent OTU classifications without a needing a reference database. We have clarified this throughout the text.}

The authors need to situate their methods within the universe of methods commonly used in this area. Closed-reference OTUs and ASVs are widely-used alternative approaches to 16S sequencing data analysis that are already used for ML classification. How does the hybrid approach proposed here of de novo OTUs plus closed-reference OTU assignment compare?

\textcolor{blue}{We thank the reviewer for this suggestion and agree that this strengthens the paper. We added the following methods: reference based clustring using OptiFit and the GreenGenes reference, \*de novo\* clustering using VSEARCH, and reference-based clustering using VSEARCH and the GreenGenes reference. See lines 47 - 64 and figure 1 for workflow descriptions and lines 87 - 98 and Figure 2 for the results.}

Is OptiFit only usable with OptiClust as the precursor?

\textcolor{blue}{OptiFit is not dependent on use of OptiClust as the precurson, you could use any algorithm to cluster the reference and then use OptiFit to fit additional data.}

What is the relevance of the MCC evaluation? What if an alternative method had lower MCC but better pred

\textcolor{blue}{The MCC score is a measure of OTU cluster quality based on the similarity of sequences and whether they are appropriately clustered together or not. Since we added additional methods, we can see that some methods do have lower MCC but equal model performance. This likely indicates that the model depends on well clustered OTUs. We've added some discussion on this to the paper (lines: ).}

### 2023-04-13 submission/manuscript.Rmd
first round revisions after pat comments

---
title: "**Streamlined implementation of a machine learning model to classify screen relevant neoplasia using reference-based OTU clustering**"
output:
  rmarkdown::pdf_document:
    keep_tex: yes
geometry: margin=1.0in
font-size: 11pt
header-includes:
  - \usepackage{helvet}
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage[left]{lineno}
  - \usepackage{multirow}
  - \usepackage[none]{hyphenat}
editor_options: 
  markdown: 
    wrap: 72
bibliography: references.bib
csl: mbio.csl
link-citations: true
---

@@

\vspace{5mm}

@@

Machine learning classification of disease using the gut microbiome
often relies on clustering 16S rRNA gene sequences into operational
taxonomic units (OTUs) to quantify microbial composition. The abundance
of each OTU is then used to train a classification model. The standard
de novo approach to clustering sequences into OTUs leverages the
similarity of the sequences to each other rather than to a reference
database. However, such an approach depends on the sequences in the
dataset and therefore OTU assignments can change if new data are added.
This lack of stability complicates classification because in order to
use the model to classify additional samples, the new sequences must be
reclustered with the old data and the model must be retrained with the
new OTU assignments. The new reference-based clustering algorithm,
OptiFit, addresses this issue by fitting new sequences into existing
OTUs. While OptiFit can produce high quality OTU clusters, it is unclear
whether this method for fitting new sequence data into existing OTUs
will impact the performance of classification models. We used OptiFit to
cluster additional data into existing OTU clusters and then evaluated
model performance in classifying a dataset containing samples from
patients with and without colonic screen relevant neoplasia (SRN). We
compared the performance of this model to the standard procedure of de
novo clustering all the data together. We found that both approaches
performed equally well in classifying SRNs. OptiFit can streamline the
process of classifying new samples by avoiding the need to retrain
models using reclustered sequences.

@@

There is great potential for using microbiome data to non-invasively
diagnose people. One of the challenges with using classification models
based on the relative abundance of operational taxonomic units (OTUs) is
that 16S rRNA gene sequences are assigned to OTUs based on their
similarity to other sequences in a dataset. If data are generated from
new patients seeking a diagnosis, then a standard approach requires
reassigning sequences to OTUs and retraining the classification model.
Yet there is a desire to have a single, validated model that can be
widely deployed. To overcome this obstacle, we applied the OptiFit
clustering algorithm which fits new sequence data to existing OTUs and
allows for the reuse of models. A random forest machine learning model
implemented using OptiFit performed as well as the traditional
reassignment and retrain approach. This result indicates that there is
potential for developing and applying machine learning models based on
OTU relative abundance data that do not require retraining.

@@

Gut community composition is useful as a resource for machine learning
classification of diseases, such as colorectal cancer [@baxter2016;
@zackular2014]. Taxonomic composition of microbial communities can be
assessed using amplicon sequencing of the 16S rRNA gene, which is the
input to classification models. Analysis of 16S rRNA gene sequence data
generally relies on similarity-based clustering of sequences into
operational taxonomic units (OTUs). The process of OTU clustering can
either be reference-based or *de novo*. The quality of OTUs generated
with reference-based clustering is generally poor compared to those
generated with *de novo* clustering @westcott2015. While *de novo*
clustering produces high-quality OTU clusters where sequences are
accurately grouped based on similarity thresholds, the resulting OTU
clusters depend on the sequences within the dataset and the addition of
new data has the potential to redefine OTU cluster composition. The
unstable nature of *de novo* OTU clustering complicates deployment of
machine learning models since integration of additional data requires
reclustering all the data and retraining the model. The ability to
integrate new data into a validated model without reclustering and
retraining could allow for the application of a single model that can
continually classify new data. Recently, Sovacool *et al.* introduced
OptiFit, a method for fitting new sequence data into existing OTUs
@sovacool2022. While OptiFit can effectively fit new sequence data to
existing OTU clusters, it is unknown if the use of OptiFit will have an
impact on classification performance. Here, we tested the ability of
OptiFit to cluster new sequence data into existing OTU clusters for the
purpose of classifying disease based on gut microbiome composition.

We compared two approaches, one using all the data to generate OTU
clusters, and the other generating *de novo* OTU clusters with a portion
of the data and then fitting the remaining sequence data to the existing
OTUs using OptiFit. In the first approach, all the 16S rRNA sequence
data was clustered into OTUs with the OptiClust algorithm in mothur
@westcott2017. The resulting abundance data was then split into training
and testing sets, where the training set was used to tune
hyperparameters and ultimately train and select the model. The model was
applied to the testing set and performance evaluated (Figure 1A).
However, this methodology requires one to regenerate the OTU clusters
and retrain the model to classify additional samples. The OptiFit
algorithm @sovacool2022 addresses this problem by enabling new sequences
to be clustered into existing OTUs. The OptiFit workflow is similar to
the OptiClust workflow, where the data was clustered into OTUs and used
to tune hyperparameters and ultimately train the model. Then, we used
OptiFit to fit sequence data of samples not part of the original dataset
into the existing OTUs, and used the same model to classify the samples
(Figure 1B). To test how the model performance compared between these
two approaches, we used a publicly available dataset of 16S rRNA gene
sequences from stool samples of healthy subjects (n = 261) as well as
subjects with SRN consisting of advanced adenoma and carcinoma (n = 229)
@baxter2016. The dataset was randomly split into an 80% train set and
20% test set. For the standard OptiClust workflow, the training and test
sets were *de novo* clustered together into OTUs, then the resulting
abundance table was split into the training and testing set. For the
OptiFit workflow, the train set was clustered *de novo* into OTUs, and
the remaining test set was fit to the OTU clusters using the OptiFit
algorithm. For both workflows, the abundance table of the train set was
used to tune hyperparameters and train a random forest model to classify
SRN. The test set was classified as either control or SRN using the
trained models. To account for variation, the dataset was randomly split
100 times and the process repeated for each of the 100 data splits. By
comparing the model performance of classifying the samples in the test
dataset between the OptiFit and OptiClust algorithms, we quantified the
impact of using OptiFit on model classification performance.

```{r mcc,include=F}
merged_mcc <- read_csv("../data/learning/summary/merged_MCC.csv")

opticlust_mcc <- merged_mcc %>% 
  filter(algorithm == "opticlust") %>% 
  pull(mcc)

optifit_avg_mcc <- merged_mcc %>% 
  filter(algorithm == "optifit" & state=="combo") %>% 
  summarise(mean_mcc = mean(mcc),
            sd_mcc = sd(mcc))
```

```{r frac_mapped,include=F}
avg_frac_mapped <- read_csv("../results/tables/fracNonMapped.csv",show_col_types = F) %>% 
  mutate(avg_frac = RefSum/10000) %>% 
  summarise(mean = mean(avg_frac),
            sd = sd(avg_frac))
```

We first examined the quality of the resulting OTU clusters from the two
algorithms using the Matthews correlation coefficient (MCC). The MCC
score was quantified by examining all pairs of sequences and assessing
whether they belonged together in an OTU based on their similarity
@westcott2017. MCC scores range between negative one and one. A score of
negative one means none of the sequences in an OTU are within the
similarity threshold and any sequences within the similarity threshold
are not in an OTU together. An MCC score of zero means the sequences are
randomly clustered. An MCC score of 1 means all sequences in an OTU are
within the similarity threshold and all sequence pairs within the
similarity threshold are in the same OTU. To ensure that OptiFit is
appropriately integrating new sequence data into the existing OTUs, we
expected the MCC scores produced by the OptiClust and OptiFit workflows
to be similar. Since the data was only clustered once in the OptiClust
workflow there was only one MCC score (MCC =
`r round(opticlust_mcc,digits=3)`) while the OptiFit workflow produced
an MCC score for the OTU clusters from each data split (average MCC =
`r round(optifit_avg_mcc %>% pull(mean_mcc),digits=3)`, standard
deviation = `r round(optifit_avg_mcc %>% pull(sd_mcc),digits=3)`). As
expected, the MCC scores were similar between the two workflows. Another
metric we examined for the OptiFit workflow was the fraction of
sequences from the test set that mapped to the reference OTUs. Any
sequences that did not map to reference OTUs were eliminated therefore,
if a high percentage of reads did not map we might expect this loss of
data to negatively impact classification performance. We found that loss
of data was not an issue since on average
`r format_decimal((avg_frac_mapped %>% pull(mean))*100,digits = 1)`%
(standard deviation =
`r format_decimal(avg_frac_mapped %>% pull(sd),digits = 3)`) of
sequences in the test set mapped to the reference OTUs. These results
indicate that OptiFit performed as well as OptiClust when integrating
new sequences into the existing OTUs.

```{r performance,include=F}
performance <- read_csv("../data/learning/summary/merged_performance.csv",
                    col_types = cols(algorithm=col_character(),
                                     method=col_character(),
                                     split=col_character(),
                                     .default=col_double()))

@@

pvals <- read_csv("../results/tables/pvalues.csv",col_types = cols(p_value = col_double(),
                                                                   .default = col_character())) %>% 
  select(metric,p_value)
```

After verifying that the quality of the OTUs was consistent between
OptiClust and OptiFit, we examined the model performance for classifying
samples in the held out test dataset. To evaluate model performance, we
used the OTU relative abundances from the training data from the
OptiClust and OptiFit workflows to train a model to predict SRNs. Using
the predicted and actual diagnosis classification, we calculated the
area under the receiver operating characteristic curve (AUROC) for each
data split to quantify model performance. During cross-validation (CV)
training, the model performance was equivalent between the OptiClust and
OptiFit approaches (p-value =
`r format_decimal(pvals %>% filter(metric == "cv_metric_AUC") %>% select(p_value),2)`;
Figure 2A). The trained model was then applied to the test data
classifying samples as either control or SRN. The performance on the
test data was equivalent between the OptiClust and OptiFit approaches
(p-value =
`r format_decimal(pvals %>% filter(metric == "AUC") %>% select(p_value),2)`;
Figures 2B and 2C) indicating that new data could be fit to existing OTU
clusters using OptiFit without impacting model performance.

We tested the ability of OptiFit to integrate new data into existing
OTUs for the purpose of machine learning classification using OTU
relative abundance. A potential problem with using OptiFit is that any
sequences from the new samples that do not map to the existing OTU
clusters will be discarded, resulting in a possible loss of information.
However, we demonstrated OptiFit can be used to fit new sequence data
into existing OTU clusters and performs equally well in predicting SRN
compared to *de novo* clustering all the sequence data together. The
ability to integrate data from new samples into existing OTUs enables
the implementation of a single machine learning model. This is important
for model implementation because not all of the data needs to be
available or known at the time of model generation. Since results may
depend on the amount of data in the training set, further analysis is
needed to determine the number of samples that are necessary in training
to build a robust model capable of classifying diverse samples. A robust
machine learning model could be implemented as part of a non-invasive
and low-cost aid in diagnosing SRN and other diseases.

@@

***Dataset.*** Raw 16S rRNA gene sequence data isolated from human stool
samples was downloaded from NCBI Sequence Read Archive (accession no.
SRP062005) [@edgar2011; @baxter2016]. This dataset contains stool
samples from a total of 490 subjects. For this analysis, samples from
subjects identified in the metadata as normal, high risk normal, or
adenoma were categorized as "normal", while samples from subjects
identified as advanced adenoma or carcinoma were categorized as "screen
relevant neoplasia" (SRN). The resulting dataset consisted of 261 normal
samples and 229 SRN samples.

```{r n_test_train,include=F}
n_test_train <- read_csv("../results/tables/splitTogetherFrequency.csv") %>% 
  mutate(n_test = rowSums(across(starts_with("split_")))) %>% 
  select(Group,n_test) %>% 
  mutate(n_train = 100 - n_test) %>% 
  summarize(avg_test = mean(n_test),
            sd_test = sd(n_test),
            avg_train = mean(n_train),
            sd_train = sd(n_train))
```

***Data Processing.*** The full dataset was preprocessed with mothur
(v1.47) @schloss2009 to join forward and reverse reads, merge duplicate
reads, align to the SILVA reference database (v132) @quast2013,
precluster, remove chimeras with UCHIME @edgar2011, assign taxonomy, and
remove non-bacterial reads following the Schloss Lab MiSeq standard
operating procedure described on the mothur website
(<https://mothur.org/wiki/miseq_sop/>). 100 splits of the 490 samples
were generated where 80% of the samples (392 samples) were randomly
assigned to the training set and the remaining 20% (98 samples) were
assigned to the test set. Using 100 splits of the data accounts for the
variation that may be observed depending on the samples that are in the
training or test sets. Each sample was in the training set an average of
`r n_test_train %>% pull(avg_train)` times (standard deviation =
`r format_decimal(n_test_train %>% pull(sd_train),digits = 1)`) and the
test set an average of `r n_test_train %>% pull(avg_test)` times
(standard deviation =
`r format_decimal(n_test_train %>% pull(sd_test),digits = 1)`).

The data was processed through two workflows. In the control workflow,
all the data was clustered together with OptiClust @westcott2017 to
generate OTUs and the resulting abundance tables were split into the
training and testing sets. In the experimental workflow, the
preprocessed data was split into the training and testing sets. The
training set was clustered into OTUs using OptiClust, then the test set
was fit to the OTUs of the training set using the OptiFit algorithm
@sovacool2022. The OptiFit algorithm was run with method open so that
any sequences that did not map to the existing OTU clusters would form
new OTUs. For both pathways, the shared files were sub-sampled to 10,000
reads per sample.

@@

***Code Availability.*** The analysis workflow was implemented in
Snakemake @koster2012 . Scripts for analysis were written in R @R2020
and GNU bash @GNUbash. The software used includes mothur v1.47.0
@schloss2009, RStudio @RStudio2019, the Tidyverse metapackage
@wickham2019, R Markdown @xie_r_2018, the SRA toolkit
@noauthor_sra-tools_nodate, and conda @noauthor_anaconda_2016. The
complete workflow and supporting files required to reproduce this study
are available at:
<https://github.com/SchlossLab/Armour_OptiFitGLNE_XXXX_2021>

## Acknowledgements

@@

![](../exploratory/figures/figure1_a.pdf)

![](../exploratory/figures/figure1_b.pdf)

**Figure 1: Workflows.** **A)** OptiClust workflow: The full dataset was
clustered into OTUs using the OptiClust algorithm in mothur. The data
was then split into two sets where 80% of the samples were assigned to
the training set and 20% to the testing set. The training set was
preprocessed with mikropml to normalize values (scale and center),
collapse correlated features, and remove features with zero variance.
Using mikropml, the training set was split into train and validate sets
to compare results using different hyperparameter settings. The highest
performing hyperparameter setting was then used to train the model with
the full training set. The preprocessing scale from the training set was
applied to the test dataset, then the trained model was used to classify
the samples in the test set. Based on the actual classification and
predicted classification, the area under the receiver operating
characteristic curve (AUROC) was calculated to summarize model
performance. The entire process was repeated 100 times to account for
variability depending on the split of the data resulting in a total of
100 AUROC values summarizing the performance of the standard OptiClust
workflow. **B)** OptiFit workflow: The dataset was first split into two
sets where 80% of the samples were assigned to the training set and 20%
to the testing set. The training set was then clustered into OTUs using
the OptiClust algorithm in mothur. The resulting abundance data was
preprocessed with mikropml to normalize values (scale/center), collapse
correlated features, and remove features with zero-variance. Using
mikropml, the training set was split into train and validate sets to
compare results using different hyperparameter settings. The highest
performing hyperparameter setting was then used to train the model with
the full training set. The OptiFit algorithm in mothur was used to
cluster the held out testing dataset using the OTUs of the training set
as a reference. The preprocessing scale from the training set was
applied to the test dataset, then the trained model was used to classify
the samples in the test set. Based on the actual classification and
predicted classification, the area under the receiver operating
characteristic curve (AUROC) was calculated to summarize model
performance. The entire process was repeated 100 times to account for
variability depending on the split of the data resulting in a total of
100 AUROC values summarizing the performance of the new OptiFit
workflow.

>>>>

---
title: "**Machine learning classification using self-reference-based OTU clustering**"
output:
  rmarkdown::pdf_document:
    keep_tex: yes
geometry: margin=1.0in
font-size: 11pt
header-includes:
  - \usepackage{helvet}
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage[left]{lineno}
  - \usepackage{multirow}
  - \usepackage[none]{hyphenat}
editor_options: 
  markdown: 
    wrap: 72
bibliography: references.bib
csl: mbio.csl
link-citations: true
---

@@

@@

Machine learning classification using the gut microbiome relies on
assigning 16S rRNA gene sequences into operational taxonomic units
(OTUs) to quantify microbial composition. OTU abundances are then used
to train a classification model that can be applied to classify new
samples. The standard approaches to clustering sequences include
reference-based and *de novo* clustering. Reference-based clustering
requires a well-curated reference database that may not exist for all
systems. *De novo* clustering tends to produce higher quality OTU
assignments than reference-based, but clusters depend on the sequences
in the dataset and therefore OTU assignments will change when new
samples are sequenced. This lack of stability complicates machine
learning classification since new sequences must be reclustered with the
old data and the model retrained with the new OTU assignments. The
OptiFit algorithm addresses these issues by fitting new sequences into
existing OTUs. While OptiFit produces high quality OTU clusters, it is
unclear whether this method for fitting new sequence data into existing
OTUs will impact the performance of classification models trained with
the older data. We used OptiFit to cluster sequences into existing OTUs
and evaluated model performance in classifying a dataset containing
samples from patients with and without colonic screen relevant neoplasia
(SRN). We compared the performance of this model to standard methods
including *de novo* and database-reference-based clustering. We found
that using OptiFit performed as well or better in classifying SRNs.
OptiFit can streamline the process of classifying new samples by
avoiding the need to retrain models using reclustered sequences.

@@

There is great potential for using microbiome data to aid in diagnosis.
A challenge with OTU-based classification models is that 16S rRNA gene
sequences are often assigned to OTUs based on similarity to other
sequences in the dataset. If data are generated from new patients, the
old and new sequences must all be reassigned to OTUs and the
classification model retrained. Yet there is a desire to have a single,
validated model that can be widely deployed. To overcome this obstacle,
we applied the OptiFit clustering algorithm to fit new sequence data to
existing OTUs allowing for reuse of the model. A random forest model
implemented using OptiFit performed as well as the traditional reassign
and retrain approach. This result shows that it is possible to train and
apply machine learning models based on OTU relative abundance data that
do not require retraining or the use of a reference database.

@@

There is increasing evidence for an association between the composition of the gut microbiome and a variety of diseases, such as crohn's disease and colorectal cancer. The signatures that associate with disease have shown potential to aid in diagnosis of disease throught gut microbiome profiling and machine learning. Taxonomic composition of microbial
communities can be assessed using amplicon sequencing of the 16S rRNA
gene, which is the input to classification models. Analysis of 16S rRNA
gene sequence data generally relies on assigning sequences into
operational taxonomic units (OTUs). The process of OTU clustering can
either be reference-based or *de novo*. The quality of OTUs generated
with reference-based clustering is generally poor compared to those
generated with *de novo* clustering @westcott2015. While *de novo*
clustering produces high-quality OTU clusters where sequences are
accurately grouped based on similarity thresholds, the resulting OTU
clusters depend on the sequences within the dataset and the addition of
new data has the potential to redefine OTU cluster composition. The
unstable nature of *de novo* OTU clustering complicates deployment of
machine learning models since integration of additional data requires
reclustering all the data and retraining the model. The ability to
integrate new data into a validated model without reclustering and
retraining could allow for the application of a single model that can
continually classify new data. Recently, Sovacool *et al.* introduced
OptiFit, a method for fitting new sequence data into existing OTUs
@sovacool2022. While OptiFit can effectively fit new sequence data to
existing OTU clusters, it is unknown if the use of OptiFit will have an
impact on classification performance. Here, we tested the ability of
OptiFit to cluster new sequence data into existing OTU clusters for the
purpose of classifying disease based on gut microbiome composition.

We compared the ability of several approaches for assigning 16S rRNA
gene seequences to OTUs including, *de novo*, database-reference-based,
and self-reference-based with OptiFit. To test how the model performance
compared between these approaches, we used a publicly available dataset
of 16S rRNA gene sequences from stool samples of healthy subjects (n =
226) as well as subjects with screen-relevant neoplasia (SRN) consisting
of advanced adenoma and carcinoma (n = 229) @baxter2016. For the *de
novo* workflows, all the 16S rRNA sequence data was clustered into OTUs.
The OTU clustering was conducted using two common algorithms: 1) the
OptiClust algorithm in mothur @westcott2017 and 2) the VSEARCH algorithm
used in QIIME2 [@rognes2016; @bolyen2019]. For both algorithms, the
resulting abundance data was then split into training and testing sets,
where the training set was used to tune hyperparameters and ultimately
train and select the model. The model was applied to the testing set and
performance evaluated (Figure 1A). We also conducted reference-based OTU
clustering using OptiFit to fit the sequence data into OTUs based on the
greengenes reference database. To compare with another commonly used
method, we also used the VSEARCH algorithm to fit the sequence data to
the greengenes reference (Figure 1B). In the OptiFit self-reference
workflow, the data was split into a training and a testing set. The
training set was clustered into OTUs and used to train a classification
model. The OptiFit algorithm was used to fit sequence data of samples
not part of the original dataset into the existing OTUs, and used the
same model to classify the samples (Figure 1C). For each of the
workflows the process was repeated for 100 random splits of the data to
account for variation caused by the choice of the random number
generator seed.

```{r mcc, include=F}
merged_mcc <- read_csv("../results/ml/summary/merged_mcc.csv")

opticlust_denovo_mcc <- merged_mcc %>% 
  filter(algorithm == "opticlust_denovo") %>% 
  pull(mcc) %>% round(digits=3)

optifit_self_avg_mcc <- merged_mcc %>% 
  filter(algorithm == "optifit_self" & state=="combo") %>% 
  summarise(mean_mcc = mean(mcc),
            sd_mcc = sd(mcc)) %>% 
  round(digits=3)

optifit_gg_mcc <- merged_mcc %>% 
  filter(algorithm == "optifit_gg") %>% 
  pull(mcc) %>% round(digits=3)

vsearch_denovo_mcc <- merged_mcc %>% 
  filter(algorithm == "vsearch_denovo") %>% 
  pull(mcc) %>% round(digits=3)

vsearch_gg_mcc <- merged_mcc %>% 
  filter(algorithm == "vsearch_gg") %>% 
  pull(mcc) %>% round(digits=3)
```

```{r frac_mapped, include=F}
counts <- read_csv("../results/tables/counts.csv")

avg_frac_mapped <- counts %>%
  mutate(frac_mapped = count/10000 * 100) %>%
  group_by(methods,set) %>%
  summarize(mean_frac = mean(frac_mapped),
            sd_frac = sd(frac_mapped),
            .groups="drop") %>% 
  filter(set == "test") 
```

We first examined the quality of the resulting OTU clusters from each
method using the Matthews correlation coefficient (MCC). MCC is a metric
used to measure OTU cluster quality based on the similarity of all pairs
of sequences and whether they are appropriately clustered or not
@westcott2015. MCC scores range between negative one and one, and
measure how well clustering assignment correlates with the distance
between sequences. To ensure that OptiFit appropriately integrated new
sequence data into the existing OTUs, we expected the MCC scores
produced by the OptiFit workflow to be similar to that of *de novo*
clustering using the OptiClust algorithm. In the OptiFit workflow the
test data was fit to the clustered training data for each of the 100
data splits resulting in an MCC score for each split of the data. In the
remaining workflows, the data was only clustered once and then split
into the training and testing sets resulting in a single MCC score for
each method. Indeed, the MCC scores were similar between the OptiClust
*de novo* (MCC = `r opticlust_denovo_mcc`) and OptiFit self-reference
workflows (average MCC = `r optifit_self_avg_mcc %>% pull(mean_mcc)`,
standard deviation = `r optifit_self_avg_mcc %>% pull(sd_mcc)`).
Previously, we observed that *de novo* clustering tends to produce
higher MCC scores than reference-based clustering @sovacool2022.
Consistent with prior findings the reference-based methods produced
lower MCC scores (OptiFit Greengenes MCC = `r optifit_gg_mcc`; VSEARCH
Greengenes MCC = `r vsearch_gg_mcc`) than the *de novo* methods with the
same algorithm (OptiClust *de novo* MCC = `r opticlust_denovo_mcc`;
VSEARCH *de novo* MCC = `r vsearch_denovo_mcc`). Another metric we
examined for the OptiFit workflow was the fraction of sequences from the
test set that mapped to the reference OTUs. Since sequences that did not
map to reference OTUs were eliminated, if a high percentage of reads did
not map to an OTU we expected this loss of data to negatively impact
classification performance. We found that loss of data was not an issue
since on average
`r avg_frac_mapped %>% filter(methods == "optifit_self", set == "test") %>% pull(mean_frac) %>% round(digits=1)`%
(standard deviation =
`r avg_frac_mapped %>% filter(methods == "optifit_self", set == "test") %>% pull(sd_frac) %>% round(digits=2)`%)
of sequences in the test set mapped to the reference OTUs. This number
is higher than the average fraction of reads mapped in the OptiFit
Greengenes workflow (
`r avg_frac_mapped %>% filter(methods == "optifit_gg",set == "test") %>% pull(mean_frac) %>% round(digits=1)`%
+/-
`r avg_frac_mapped %>% filter(methods == "optifit_gg", set == "test") %>% pull(sd_frac) %>% round(digits=1)`).
These results indicate that the OptiFit self-reference method performed
as well as the OptiClust *de novo* method and better than using an
external database.

```{r}
performance <- read_csv("../results/ml/summary/merged_performance.csv",
                    col_types = cols(algorithm=col_character(),
                                     method=col_character(),
                                     split=col_character(),
                                     .default=col_double()))

@@

pvals <- read_csv("../results/tables/pvalues.csv",col_types = cols(p_value = col_double(),
                                                                   .default = col_character()))

get_pval <- function(pvals,met,g1,g2){
  pvals %>% 
    filter(metric == met) %>% 
    filter(group1 == g1 & group2 == g2 | group1 == g2 & group2 == g1) %>% 
    pull(p_value) %>% round(digits = 3)
}
```

We next assessed model performance using OTU relative abundances from
the training data from the workflows to train a model to predict SRNs
and used the model on the held out data. Using the predicted and actual
diagnosis classification, we calculated the area under the receiver
operating characteristic curve (AUROC) for each data split. During
cross-validation (CV) training, the model performance was equivalent
between OptFit self-reference and OptiClust *de novo* (p-value =
`r get_pval(pvals, "cv_metric_AUC", "opticlust_denovo", "optifit_self")`;
Figure 2A) while while performance for both VSEARCH methods was lower
than the OptiClust *de novo*, OptiFit self, and OptiFit Greengenes
methods (p-values \< 0.05). The trained model was then applied to the
test data classifying samples as either control or SRN. Both VSEARCH
methods perform slightly worse than the OptiClust \*de novo\* method
(both p-values \< 0.05). However the performance on the test data was
equivalent between the OptiClust *de novo*, OptiFit Greengenes, and
OptiFit self-reference approaches (p-value \> 0.05; Figures 2B and 2C).
These results indicate that new data could be fit to existing OTU
clusters using OptiFit without impacting model performance.

We tested the ability of OptiFit to integrate new data into existing
OTUs for the purpose of machine learning classification using OTU
relative abundance. A potential problem with using OptiFit is that any
sequences from the new samples that do not map to the existing OTU
clusters will be discarded, resulting in a possible loss of information.
However, we demonstrated that OptiFit can be used to fit new sequence
data into existing OTU clusters and it could perform as well in
predicting SRN compared to *de novo* clustering all the sequence data
together. In this instance, the performance of OptiFit was equivalent to
using a database-reference-based method despite the lower quality of the
OTU clusters in the database-reference-based approach. This likely
indicates that the sequences that are important to the model are well
characterized by the reference database. However, a less well studied
system may not be as well characterized by a reference-database which
would make the ability to utilize one's own data a reference an exciting
possiblility. The ability to integrate data from new samples into
existing OTUs enables the implementation of a single machine learning
model. This is important for model implementation because not all of the
data needs to be available or known at the time of model generation. A
robust machine learning model can be implemented as part of a
non-invasive and low-cost diagnostic for SRN and other diseases.

@@

**Dataset.** Raw 16S rRNA gene sequence data from the V4 region were
previously generated from human stool samples. Sequences were downloaded
from the NCBI Sequence Read Archive (accession no. SRP062005)
[@edgar2011; @baxter2016]. This dataset contains stool samples from 490
subjects. For this analysis, samples from subjects identified in the
metadata as normal, high risk normal, or adenoma were categorized as
"normal", while samples from subjects identified as advanced adenoma or
carcinoma were categorized as "screen relevant neoplasia" (SRN). The
resulting dataset consisted of 261 normal samples and 229 SRN samples.

```{r n_test_train,include=F}
n_test_train <- read_csv("../results/tables/train_test_frequency.csv") %>% 
  pivot_wider(names_from="train_test",values_from = "n") %>% 
  summarize(avg_test = mean(test),
            sd_test = sd(test),
            avg_train = mean(train),
            sd_train = sd(train))
```

**Data processing.** The full dataset was preprocessed with mothur
(v1.47) @schloss2009 to join forward and reverse reads, merge duplicate
reads, align to the SILVA reference database (v132) @quast2013,
precluster, remove chimeras with UCHIME @edgar2011, assign taxonomy, and
remove non-bacterial reads following the Schloss Lab MiSeq standard
operating procedure described on the mothur website
(<https://mothur.org/wiki/miseq_sop/>). 100 splits of the 490 samples
were generated where 80% of the samples (392 samples) were randomly
assigned to the training set and the remaining 20% (98 samples) were
assigned to the test set. Using 100 splits of the data accounts for the
variation that may be observed depending on the samples that are in the
training or test sets. Each sample was in the training set an average of
`r n_test_train %>% pull(avg_train)` times (standard deviation =
`r format_decimal(n_test_train %>% pull(sd_train),digits = 1)`) and the
test set an average of `r n_test_train %>% pull(avg_test)` times
(standard deviation =
`r format_decimal(n_test_train %>% pull(sd_test),digits = 1)`\`).

***Reference-based workflows.***

1.  OptiFit Self: The preprocess data was split into the training and
    testing sets. The training set was clustered into OTUs using
    OptiClust, then the test set was fit to the OTUs of the training set
    using the OptiFit algorithm @sovacool2022. The OptiFit algorithm
    was run with method open so that any sequences that did not map to
    the existing OTU clusters would form new OTUs. The data was then
    subsampled to 10,000 reads and any novel OTUs from the test set were
    removed. This process was repeated for each of the 100 splits
    resulting in 100 training and testing datasets.
2.  OptiFit Greengenes: Reference sequences from the Greengenes database
    v13_8\_99 @desantis2006 were downloaded and processed with mothur by
    trimming to the V4 region and clustered *de novo* with OptiClust
    @westcott2017. The preprocessed data was fit to the clustered
    reference data using OptiFit with the method open to allow any
    sequences that did not map to the existing reference clusters would
    form new OTUs. The data was then subsampled to 10,000 reads and any
    novel OTUs from the test set were removed. The data set was then
    split into two sets where 80% of the samples were assigned to the
    training set and 20% to the testing set. This process was repeated
    for each of the 100 splits resulting in 100 training and testing
    datasets.
3.  VSEARCH Greengenes: Preprocessed data was clustered using VSEARCH
    v2.15.2 @rognes2016 directly to unprocessed Greengenes 97% OTU
    reference alignment consistent with how VSEARCH is typically used by
    the QIIME2 software for reference-based clustering @bolyen2019. The
    data was then subsampled to 10,000 reads and any novel OTUs from the
    test set were removed. The data set was then split into two sets
    where 80% of the samples were assigned to the training set and 20%
    to the testing set. This process was repeated for each of the 100
    splits resulting in 100 training and testing datasets.

***De novo workflows.***

4.  OptiClust *de novo*: All the preprocessed data was clustered
    together with OptiClust @westcott2017 to generate OTUs. The data was
    subsampled to 10,000 reads per sample and the resulting abundance
    tables were split into the training and testing sets. The process
    was repeated for each of the 100 splits resulting in 100 training
    and testing datasets.

5.  VSEARCH *de novo*: All the preprocessed data was clustered using
    VSEARCH v2.15.2 @rognes2016 with 97% identity and then subsampled to
    10,000 reads per sample. The process was repeated for each of the
    100 splits resulting in 100 training and testing datasets for both
    workflows.

@@

***Code Availability.***

The analysis workflow was implemented in Snakemake @koster2012. Scripts
for analysis were written in R @R2020 and GNU bash @GNUbash. The
software used includes mothur v1.47.0 @schloss2009, VSEARCH v2.15.2
@rognes2016, RStudio @RStudio2019, the Tidyverse metapackage
@wickham2019, R Markdown @xie_r_2018, the SRA toolkit
@noauthor_sra-tools_nodate, and conda @noauthor_anaconda_2016. The
complete workflow and supporting files required to reproduce this study
are available at:
<https://github.com/SchlossLab/Armour_OptiFitGLNE_mBio_2023>

## Acknowledgments

@@

![](./figures/Figure1.png)

**Figure1: Overview of clustering workflows.** The *de novo* and
database-reference-based workflows were conducted using two approaches:
OptiClust with mothur and VSEARCH as is used in the QIIME pipeline.
